A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-04-09 paper/paper-revision.Rnw
Finish abstract, changes throughout intro and methods

\begin{abstract}
To assess random match probability at the level of class characteristics, one could imagine sampling shoes from a local population and calculating the proportion of shoes "similar" to a given print. Our proposed implemention of such a procedure involves 1) a scanner to gather images of shoes and 2) software to automatically label images within a feature space and assess the images for similarity among those features. In this paper, we discuss an algorithm for automatically labeling shoe outsole images using a feature space similar to what examiners use: a set of features based on geometric shapes such as circles, stars, and quadrilaterals. The automatic labeling algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant improvement to the current manual classification of footwear tread patterns, and could be integrated into current matching databases to reduce the amount of time examiners spend on data entry.
\end{abstract}

@@

\section{Introduction}

In the United States, forensic analysis of shoe print evidence in criminal cases is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. In the cases where individualizing characteristics are not used, it may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristics; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc).

From a statistical perspective, random match probability could theoretically be calculated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. In practice, one potential method of sampling could be a low-profile scanner placed in a high traffic area. Such a scanner could take \mt{impressions or} pictures of shoes walking past, and those scans could then be used to create a database of shoes relevant to the local area.

However, studies by \citep{parent_sandy_significance_nodate} and \citep{benedict_geographical_2014} demonstrate that even a couple hundred shoe prints constitute an unweildy amount of data that then requires \fix{[a large amount of data processing]}. Thus, in parallel with the development of such a scanner, it is necessary to develop a way to process large amounts of data automatically. \mt{The primary focus of this paper is the development of one such method for large-scale automatic data processing, where "data processing" refers to assessing} the similarity of the shoe tread patterns in the collected data.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations that may experience inclement weather or other non-controlled conditions. \mt{Thus, we are working to develop a scanner using photo technology. This scanner would take pictures of shoe outsoles, much like images that are currently found on the websites of online shoe retailers \autoref{fig:prototype-scanner}. Although the images from the scanner are more degraded in some ways than those found from online retailers, the fundamental requirements to analyze such data would be similar.} Therefore, \fix{we move forward by considering how to assess similarity of tread patterns in images of outsoles, in particular.} More specifically, this assessment is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

\fix{How to make the scanner images a figure instead of a table?}

@@

\fix{Rough idea: A reasonable criticism of this approach is that images from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed earlier in this paper. This critique is especially fair considering that CNNs, and neural networks in general, have been shown to perform poorly on test data when it does not resemble the training data (find citation). However, this research is so new that we need to start somewhere. We need to work the kinks out of recognizing characteristics that examiners use. We can do that on real data once the scanner has been built and once we know how well our classification works. AKA find the pitfalls for clean images rather than introduce too many points of failure at once. Once this work shows that CNNs can sufficiently distinguish outsole patterns/features, we can justify the effort to modify this approach for real/degraded images.}

>>>>

\begin{abstract}
To assess random match probability at the level of class characteristics, one could imagine sampling shoes from a local population and calculating the proportion of shoes "similar" to a given print. Our proposed implemention of such a procedure involves 1) a scanner to gather images of shoes from a local population and 2) software to automatically label images within a feature space and assess the images for similarity among those features. In this paper, we discuss an algorithm for automatically labeling shoe outsole images using a classification scheme similar to what examiners use: a set of features based on geometric shapes such as circles, stars, and quadrilaterals. The automatic labeling algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant step toward making tractable the computation of random match probabilities of tread patterns, and could potentially be modified to automate other types of footwear examinations based on class characteristics.
\end{abstract}

@@

\section{Introduction}\label{sec:Introduction}

In the United States, forensic analysis of shoe print evidence is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. In the cases where individualizing characteristics are not used, it may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristics; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc).

From a statistical perspective, random match probability could theoretically be calculated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. In practice, one potential method of sampling could be a low-profile scanner placed in a high traffic area. Such a scanner could take impressions or pictures of shoes walking past, and those scans could then be used to create a database of shoes relevant to the local area.

However, a number of footwear studies (such as those by \citep{parent_sandy_significance_nodate} and \citep{benedict_geographical_2014}) have demonstrated that even a couple hundred shoe prints constitute an unweildy amount of data that require a large amount of time and effort to process and analyze. Thus, in parallel with the development of the aforementioned scanner, it is necessary to develop a way to process large amounts of data automatically. The primary focus of this paper is the development of one such method for large-scale automatic data processing, where "data processing" refers to assessing the similarity of the shoe tread patterns in the collected data.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations, which may experience inclement weather or other non-controlled conditions. Thus, we are working to develop a scanner that uses cameras to take pictures of shoe outsoles, much like images that are currently found on the websites of online shoe retailers \autoref{fig:prototype-scanner}. Although the images from the scanner are more degraded in some ways than images found from online retailers, the fundamental requirements to analyze such data would be similar. Therefore, \fix{we move forward by considering how to assess similarity of tread patterns in images of outsoles, in particular.} More specifically, this assessment is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

@@

\mt{A reasonable critique of this approach is that images collected from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed in \autoref{sec:Introduction}, and thus the CNN described in this paper is not immediately ready to process real degraded images. This concern is especially fair when considering that machine learning methods tend to perform poorly when applied to test data that do not resemble the training data. However, the proposed automation methods for this approach are quite novel in the domain of footwear examination, and a fully functioning scanner does not yet exist to collect large amounts of degraded images. Thus, the goal of this work is to evaluate the proposed classification scheme and determine whether CNNS can sufficiently distinguish features found on outsoles. Addressing these questions with clean retailer images provides both justification and guidance for modifying this approach for real degraded images.}
